---
title: "Simple spatial effort and density"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

A simple illustration of how to obtain measures of effort, density and abundance with minimal coding based on Swedish brown bear monitoring data.

Load data (`n` is number of captures) and a grid to predict on.
```{r}
library(tidyverse)
data <- read_csv("bear_captures.csv") %>% 
  filter(year == 2020)
data
predict_grid <- read_csv("grid_ZY.csv") # Just a spatial grid for visualisations
```

Fit a spatial GAM (thin-plate spline) with number of captures as response, including some ad-hoc fixes to account for 
the lack of a zero-truncated Poisson family in `mgcv`.

```{r}
initial_fit <- mgcv::gam((n - 1) ~ s(lon, lat, k = 100), 
                         family = "poisson", 
                         data = data)
mu_tilde <- predict(initial_fit, 
                    newdata = predict_grid,
                    type = "response") + 1
fit_mu <- Vectorize(function(xbar){
  # Fits a zero-truncated Poisson based on observed mean
  uniroot(f = function(mu) mu / (1 - exp(-mu)) - max(xbar, 1.05), 
          interval = c(0.01, 100))[["root"]]
})
mu <- fit_mu(mu_tilde) # Mean number of captures
q <- 1 - exp(-mu) # Probability of capture
```

Fit a spatial intensity using a kernel density estimate.

```{r}
H <- ks::Hpi.diag(cbind(data$lon, data$lat))
lambda_q <- ks::kde(cbind(data$lon, data$lat), H = H, 
                    eval.points = predict_grid)[["estimate"]] * nrow(data)
lambda <- lambda_q / q # Animal density
```

Illustrate search effort (as mean number of captures).


```{r}
predict_grid %>% 
  ggplot(aes(x = lon, y = lat)) + 
  geom_contour_filled(aes(z = mu)) + 
  geom_point(data = data, color = "white", size = .01) +
  theme_void()
```

Illustrate probability of capture.

```{r}
predict_grid %>% 
  ggplot(aes(x = lon, y = lat)) + 
  geom_contour_filled(aes(z = q)) +
  geom_point(data = data, color = "white", size = .01) + 
  theme_void()
```

Animal density (individuals per km2).

```{r}
predict_grid %>% 
  ggplot(aes(x = lon, y = lat)) + 
  geom_contour_filled(aes(z = lambda * 10^6)) + theme_void() +
  geom_point(data = data, color = "white", size = .01) +
  theme_void()
```

Estimating abundance.

```{r}
mu_tilde_obs <- predict(initial_fit, type = "response") + 1
mu_obs <- fit_mu(mu_tilde_obs)
q_obs <- 1 - exp(-mu_obs)
sum(1 / q_obs) %>% round()
```



